Technical Dimensions of Programming Systems

### 2022-09-27 prog-2022/changelog.tex
New cover letter, ensure nbsp pre cites

We graciously thank the reviewers for their detailed feedback. We've
made many minor changes to the paper and several major changes to both
structure and content.

Two issues particularly stood out:

@@

All three reviewers requested clarification on what constitutes a
``system'', mentioning IDEs amongst other things as having unclear
membership. Unfortunately, we truly don't have a definition behind the
scenes which we're using to include or exclude examples. It's more of a
``family resemblance'' between things that generalize ``programming
languages'' to ``software which lets one program, yet doesn't have to be
a language''. Nevertheless, we thought it might be worth adding a new
classification scheme for the examples we mention, which should get
across the key ``reference classes'' from which we're drawing. In sum,
these are: things that are like operating systems (O-type), things that
are like domain-specific applications (A-type) and the systems that are
based around a programming language (L-type). We also went into detail
about how a programming language (Java) can be interpreted as a system
in several different ways. These were all additions to Section 3.

\paragraph{But how do you use it?}

Two reviewers were concerned that it's hard to see how to use the
framework. We've rectified this by including a new, concrete
``Discussion'' section with two halves: one using the framework for
analysis, the other for synthesis. Specifically, we analyze Dark, a
recent programming system, along many of the dimensions. Then, we plot
two dimensions (self-sustainability vs.~notational diversity) against
each other to reveal an unexplored gap. The methodology we used to turn
qualitative dimensions into numbers is given in a new Appendix. Various
points earlier in the paper (e.g.~the Abstract and Figure 1) have been
updated to set up for this new section.

Other changes:

\begin{itemize}
\tightlist
\item
  We've significantly reworked Section 3 to be more to-the-point about
  the example systems and less of a historical narrative.
\item
  We've argued for why we think ``sociability'' belongs as a technical
  dimension in the relevant section.
\item
  We've reorganized and re-worded parts of Section 2. We've also made
  the part about how the framework relates to Cognitive Dimensions in
  Section 2, under ``Already-known characteristics'', more explicit.
\item
  We've clarified some of the points about Direct Manipulation
  (primitive actions) and Immediate Feedback (pure code previews).
\item
  Under ``Addressing and Externalizability'', we've made it explicit
  that CSS provides Additive Authoring for the Web programming system.
\item
  To make room for the ``Discussion'' section, we've had to move more
  material to the existing Appendix with the bulk of the dimensions.
  Now, the main body only contains the Customizability dimensions.
\item
  We've elevated the singleton Automation cluster to a doublet by adding
  a Factoring of Complexity dimension.
\end{itemize}

>>>>

We thank the reviewers for their continued attention and feedback. The
major changes we have made are as follows.

@@

We worked out a sensible explicit definition of what a programming
system is and included it as early as possible in the paper.

\paragraph{L/O/A typology.}

We intended the three-way classification in section~3 merely to provide
some rhyme and reason as to where our example systems were coming from.
We did not intend it as a rigid scheme where membership in a category
means we are making some profound or bold claim. The L/O/A-type naming
seems to have misled readers into anticipating the latter, so we have
relaxed the naming to ``language-based'', ``OS-like'' and
``application-focused''. We did refer back to this scheme in one
specific context, in section~5.2, noting that most OS-likes (Unix, Lisp
and Smalltalk) cluster in a region to the right.

\paragraph{Dark \& evaluation.}

Section~5 has been clarified as an evaluation of the framework by means
of the two example usages. We've included a summary table for the
dimensional analysis of Dark in section~5.1 and annotated the screenshot
for convenience. We appreciate the reviewer's desire for examples of the
specific interactions, but we thought it'd be pointless to write down
something already explained in one of Dark's tutorial videos, so we
cited the latter instead.

\paragraph{References to strengthen claims.}

Our statement about ``cool/interesting'' was entirely from experience,
so we relaxed it to something we could more easily support with
references. These can be seen in a footnote with our remarks linking
them to the point we're trying to make. Elsewhere we have addressed this
by deleting the claim (e.g.~for ``holistic view'') or adding references,
for example to the various example systems like Java and Jupyter. We're
afraid we don't understand the request to put explicit examples in the
section on venues for publishing programming systems research; we'd
already listed UIST, VL/HCC, LIVE and PX. If this was about specific
publications instead, we've now referred to some in the footnote on page
3.

\paragraph{Socio-Technical dimensions?}

The fact of the matter is that almost all of our 22 dimensions are
technical, with the possible exception of Learnability and Sociability
being socio-technical, and some remarks in Integrity vs.~Openness. On
this basis, we strongly disagree with calling the contribution
``socio-technical'' dimensions or similar, as this would give a false
expectation to readers. We appeal to the reviewer to permit us to keep
the name ``technical''---we put a lot of effort into thinking of names
early on, but ``technical dimensions'' was the best out of perhaps only
two viable choices. If this is still unacceptable, we will fall back on
the other name, ``design dimensions'', but if this is still rejected we
would need some help determining an alternative.

\paragraph{On the new length of the paper.}

Following approval from the editors, this revision includes the full
catalogue of dimensions in the main body of the paper. While we
acknowledge this makes it very long, we do not intend reviewers or
readers to read this section from start to finish. Instead, it is meant
as a reference to be used as needed, for example to get further detail
on the dimensions we reference later for the Dark programming system.
This lets us present the material as we originally intended and
addresses a reviewer concern that it was awkward to refer to appendix
material in the section on Dark. In summary, we always saw the
dimensions catalogue as belonging in the main body, but for review
purposes it may still be treated like an appendix.
